---
title: "ESM 206 Lab Week 3 Key"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(here)
library(gapminder)
```

### Part 0: Setup

- Create a new R project (does not need to be version-controlled)
- Add two subfolders: data and src
- Download two data files from GauchoSpace: jornada_lizards.csv and lizard_abb.csv, and copy them into your project data folder
- Open a new R Markdown document in your project, delete all below the setup chunk
- Attach the necessary packages in the setup chunk (tidyverse, janitor, here, gapminder). Note, you probably need to install the `gapminder` package by running `install.packages("gapminder")` in the Console

### Part 1: Tidying data continued

In this section, we'll learn how to combine two data frames with `full_join()`, and to reshape with `pivot_longer()` and `pivot_wider()` to get data into tidy format.

A. Join lizard common names with lizard data

The following data are from Jornada Basin LTER: 

- Title: Lizard pitfall trap data (LTER-II, LTER-III)
- URL: https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-jrn.210007001.36
- Citation: Lightfoot, D. 2017. Lizard pitfall trap data (LTER-II, LTER-III) ver 36. Environmental Data Initiative. https://doi.org/10.6073/pasta/ff37b4cace16a9943575f3fd7067064e (Accessed 2020-07-05).
- Lizard code list from: https://jornada.nmsu.edu/content/lizard-code-list

Read in both lizards datasets: 

- jornada_lizards.csv (pitfall trap data), which contains:
     - common name
     - site (site name)
     - sex (M = male, F = female, J = juvenile)
     - tail (B = broken, W = whole)
- lizard_abb.csv (lizard abbreviations & common names)


```{r}
# Read in lizards data
# Note: your path will differ depending on uyour project setup
lizard_names <- read_csv(here("esm206-f2021-lab3","lizard_abb.csv"))

lizard_traps <- read_csv(here("esm206-f2021-lab3","jornada_lizards.csv"))
```

Join together by the `common_name` variable with `full_join()`. See more information on `full_join()` - there are other options for more discriminating / excluding joins, but if you're not sure how they work **FULL JOIN** and deal with filtering later! 

```{r}
lizard_join <- lizard_traps %>% 
  full_join(lizard_names)
```
Notice that it will first look for matching names, and join by those. But we don't *have* to have matching names. For example, let's make a version of lizard_traps that has common_name renamed as "lizard_name":

```{r}
lizard_traps_rn <- lizard_traps %>% 
  rename(lizard_name = common_name)
```

We can then specify which column to join by: 

```{r}
lizard_join_2 <- lizard_traps_rn %>% 
  full_join(lizard_names, by = c("lizard_name" = "common_name"))
```

Refresh our memory on finding counts: 

```{r}
tail_counts <- lizard_join %>% 
  filter(common_name %in% c("Western Whiptail","Eastern Fence")) %>% 
  count(common_name, tail)

# Or, using tools we already have: 
tail_counts_2 <- lizard_join %>% 
  filter(common_name %in% c("Western Whiptail", "Eastern Fence")) %>% 
  group_by(common_name, tail) %>% 
  summarize(count = n())

tail_counts

tail_counts_2
```

B. `pivot_longer()` and `pivot_wider()`

Check out the gapminder data (reminder: why not put this `View(gapminder)` in your R Markdown code?)

Wrangle it a bit:
```{r}
n_am <- gapminder %>% 
  filter(country %in% c("United States", "Canada", "Mexico")) %>% 
  select(country, year, pop)
```
Note: these data are already in tidy format. So you probably wouldn't want to do this: 

Use `pivot_wider` to split years across multiple columns:

```{r}
n_am_wider <- n_am %>% 
  pivot_wider(names_from = year, 
              values_from = pop)
```

**Much more often** this wide format is the problem (e.g. year split up over multiple columns, but should be in a single column). We can use `pivot_longer()` to convert from wide-to-long format: 

```{r}
n_am_longer <- n_am_wider %>% 
  pivot_longer(cols = `1952`:`2007`,
               names_to = "year",
               values_to = "pop"
  ) %>% 
  mutate(year = as.numeric(year))
```

Finally, let's just do a quick ggplot reminder: 

```{r}
ggplot(n_am_longer, aes(x = year, y = pop)) +
  geom_line(aes(color = country)) +
  geom_point(aes(color = country)) +
  theme_minimal()
```

### Part 2: Writing functions in R

What is a function? A function is a named bit of code that is meant to perform specific operations or tasks with user-defined arguments. 

We've already been using tons of functions! All of our wrangling has been done with functions (`select()`, `filter()`, `mean()`, etc.) - but sometimes we might want to make our own. 

To make our own function, we need to: 

1. Name it
2. Tell it what the arguments are (choose useful names here, too!)
3. Specify the operation or tasks

The easiest way: write the equation like a normal human, then use Command + Option + X as a shortcut to put it into function notation! 

```{r}
# Creating a distance function: 
distance <- function(rate, time) {
  rate*time
}

# Then try it out with your own arguments added! 
distance(rate = 60, time = 2)
```
Example: the relationship between fish length (L) and weight (W) is estimated by: 

$W=aL^b$

A summary of parameters for different fish can be found in Ogle, D. FishR Vignette - Fish Condition and Relative Weights (2013). 

For red snapper, the parameters are estimated (from Diaz 2004) as: 

- a = 0.00001
- b = 3.076

with length (L) in cm and weight (W) in kg.

Write a function for the red snapper relationship, then estimate the weight of a red snapper that is 42 cm long.

```{r}

fish_weight <- function(a, L, b) {
  a*(L^b)
}

fish_weight(a = 0.00001, L = 42, b = 3.076)
# 0.98 kg

# Note: you can also just add arguments in the correct ORDER without specifically naming them, but I prefer being explicit. For example: 

fish_weight(0.00001, 74, 3.076)

# versus (same output): 

fish_weight(a = 0.00001, L = 42, b = 3.076)
```

Cool! Now you can make your own functions, with arguments that you name and specify! 

## End Lab 3
